A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-06 doc/popl2013/intro.tex
intro

Line 1 defines a type that is a {\em region relationship} {\tt rr[ rll, rlr, rtl, rtr]}.  Region relationships are
constrained existential types.  The {\tt tree} region relationship defines the type of a node of the tree: it has a pointer
to a list node and two pointers to the left and right subtrees. The crucial part is the {\em subregion} and {\em disjointness}
constraints that describe the relationships between the various regions.  Here {\tt rl} is the region holding all list elements,
{\tt rls} is the region holding

>>>>

Line 1 defines a type that is a {\em region relationship} {\tt rr[
rll, rlr, rtl, rtr]}.  Region relationships are constrained
existential types.  The {\tt tree} region relationship defines the
type of a node of the tree. The naming convention followed throughout
the example is that regions that begin with {\tt rl} name regions
holding list elements and regions that begin with {\tt rt} name
regions holding tree elements.  The most important part is the {\em
subregion} and {\em disjointness} constraints that describe the
relationships between the various regions.  Here {\tt rl} is the
region holding all list elements and {\tt rls} and {\tt rt} are the
regions holding just the list elements pointed to by and the tree
nodes included in the tree.  Note that {\tt rls} is required to be a
subregion of {\tt rl}, which means that all the values in region {\tt
rls} are also in region {\tt rl}.  The body of the {\tt tree} region
relationship declares the shape of a node of the tree, which includes
a pointer to a list element in {\tt rls} and pointers to two
subtrees. The region relation says that there exist four regions {\tt
rll, rlr, rtl}, and {\tt rtr} such that the left subtree has list and
tree elements that are in subregions {\tt rll} and {\tt rtl} of the
corresponding list and tree regions {\tt rls} and {\tt rt} for the
entire tree; the right subtree's list and tree elements in subregions
{\tt rlr} and {\tt rtr} are also subregions of {\tt rls} and {\tt rt}
respectively.  Furthermore, the constraints $\tt rll \ast rlr$ and
$\tt rtl \ast rtr$ guarantee that the left and right subtrees are
disjoint.

The {\tt make\_tree} task, which builds a {\tt tree} indexing a list
of integers, illustrates region {\em partitions}, another important
building block of Legion programs.  The region arguments to {\tt
make\_tree} are {\tt rl}, the region holding all list elements, {\tt
rls}, the subregion of {\tt rls} holding those list elements to be
indexed by this call to {\tt make\_tree}, and {\tt rt}, the region to
hold the tree itself.  Line 21 divides the list elements in {\tt rls}
into those that are less than and those that are greater than a split
value (which is arbitrarily chosen to be the value of the first
element of the list).  Partitioning is done by constructing a {\em
coloring} using the task {\tt color\_list}, which traverses the list
and for each list element $e$ in the subregion (called {\tt rs} in
{\tt color\_region}) either colors the element 1 or 2 (colors are
integers) depending on its relationship to the split value.  A
coloring is just a distinghuished function, built using the Legion
primitive {\tt color}, that maps elements of a region to integers.
When used in a {\tt partition} expression the coloring defines how to
populate new subregtions of the partitioned region.  On line 22, the
list elements in {\tt rls} are partitioned into two subregions {\tt
rll} and {\tt rlr} of list elements less than, and greater than or
equal to, the split value, respectively.  Another use of {\tt
partition} is seen on line 21 where the {\tt rt} region is partitioned
into subregions to hold the left {\tt rlt} and right {\tt rtr}
subtrees, respectively.  Because region {\tt rt} is empty at this
point (has no elements), the empty coloring function {\tt newcolor rt}
for the region {\tt rt} is given to the partitioning operation; the
two new subregions are also empty.  These examples illustrate that
partitioning can be used for two distinct, and equally important,
styles of computation: {\em allocate-then-partition}, in which a
global collection of data is created which is then divided up for
(usually parallel) processing, and {\em partition-then-allocate}, in
which a collection of empty but disjoint subregions is created and
then populated with data on which (usually parallel) processing will
take place.

Lines 26-27 illustrates {\em packing} a region relationship into the heap.
Regions are first-class values in Legion and can be stored in data structures.
Here a tree node is written to the {\tt root} pointer after being {\tt packed}.
Opertionally, the four existentially quantified regions {\tt rll, rlr, rtl} and {\tt rtr} are
stored with the tree value.  The type system guarantees statically that the packed value
satisfies the subregion and disjointness constraints declared on lines 3-4 at the point
the packed value is stored.  The advantage is that when the region relationship value is read 
from the heap and {\tt unpack} is applied (line 32) the newly introduced regions are already known to 
satisfy these constraints.  A key design decision in Legion, one that turns out to be
central to all the soundness results as well as the ability to perform distributed, parallel
scheduling of tasks, is that permissions and coherence annotations (which we have not yet discussed)
are {\em not} captured by region relationships.  When a region is stored in the heap it has no
permissions or coherence associated with it; these belong to the task.  When the region relationship
is read out of the heap and unpacked, the task must already hold permissions for an unpacked region 
to use it. Thus, on line 32, the task {\tt reverse\_tree} unpacks the region {\tt rtl} (among others).
The reads and writes of {\tt rtl} typecheck, however, because {\tt reverse\_tree} has read and
write permission for region {\tt rt} and $\tt rtl \leq rt}$ is a constraint of the region relationship (note
that the region {\tt rt} is not one of the existentially quantified regions!).
